\section{Introduction}

Software vulnerabilities are a critical and growing threat to modern computing systems. The National Vulnerability Database (NVD) reported over 25,000 new Common Vulnerabilities and Exposures (CVEs) in 2024 alone, highlighting the urgent need for automated and scalable vulnerability detection mechanisms. While Deep Learning-based Vulnerability Detection (DLVD) methods have demonstrated significant promise, they continue to suffer from two fundamental barriers that impede their adoption in real-world scenarios: the generalisation gap and the data silo problem.

First, state-of-the-art DLVD models exhibit a catastrophic \emph{generalisation gap} when applied across different software projects \cite{chakraborty2022deep}. For instance, a model trained on the source code of FFmpeg typically fails to reliably detect vulnerabilities in OpenSSL, even when both projects manifest the identical Common Weakness Enumeration (CWE) type. Replication studies consistently report performance drops of 50\% to 90\% in cross-project evaluation settings compared to within-project evaluations \cite{daoudi2024revisiting, ding2025vulnerability}. This degradation occurs because existing models tend to overfit to project-specific lexical patterns (e.g., API names, variable naming conventions) rather than learning the underlying structural semantics of the vulnerability.

Second, the \emph{data silo problem} prevents the aggregation of sufficient training data. Real-world vulnerability data is highly sensitive and routinely proprietary. Organisations such as financial institutions, cloud service providers, and defence contractors cannot legally or practically share their raw source code with external entities due to intellectual property concerns and privacy regulations. Consequently, a centralised DLVD model trained on one organisation's codebase cannot leverage the knowledge embedded in another organisation's repositories, severely restricting the model's exposure to diverse vulnerability manifestations.

Federated Learning (FL) has emerged as a promising paradigm to address the data silo problem by enabling collaborative model training without centralising raw data \cite{mcmahan2017communication, abadi2016deep}. However, naively applying standard FL techniques, such as FedAvg, to vulnerability detection does not resolve the generalisation gap \cite{he2021fedgraphnn}. Weight averaging across highly heterogeneous, non-IID (Independent and Identically Distributed) codebases fails to transfer structural vulnerability knowledge effectively, leading to suboptimal cross-project detection performance \cite{yamamoto2023flr, hu2024vulfl}.

The root cause of both barriers stems from the representation level: existing approaches primarily model code at the lexical or token level rather than at the structural or semantic level \cite{li2018vuldeepecker, zhou2019devign}. A buffer overflow in FFmpeg and a buffer overflow in OpenSSL are structurally identical within their Code Property Graphs (CPGs)---both typically involve a loop incrementing an index, an array access using that index, and a missing bounds check---yet they utilise completely different function names and API calls. 

To overcome these challenges, we propose \textbf{VulMorph-Fed} (Vulnerability Morphology-Guided Federated Learning), a novel federated learning framework designed to achieve robust cross-project vulnerability detection while preserving data privacy. VulMorph-Fed operates on the principle that vulnerability patterns are structurally invariant across projects but lexically heterogeneous. By extracting and federating \emph{vulnerability morphologies} rather than raw parameters or lexical features, VulMorph-Fed successfully bridges the generalisation gap without requiring centralised code sharing.

The primary contributions of this paper are:
\begin{itemize}
    \item We introduce the \textbf{Vulnerability-Critical Subgraph Abstraction (VCSA)}, an end-to-end differentiable module that extracts project-invariant structural vulnerability patterns (morphologies) by masking irrelevant edges and mapping node labels to abstract semantic types.
    \item We propose \textbf{Morphology-Conditioned Federated Prototype Aggregation (MCFPA)}, a novel FL aggregation mechanism that shares CWE-conditioned semantic prototypes instead of model parameters, significantly reducing communication costs and enabling prototype-level Differential Privacy (DP).
    \item We design \textbf{Morphology-Guided Message Passing (MGMP)} to inject global prototype signals into local Graph Neural Network (GNN) training, allowing clients to benefit from global structural knowledge across all participating projects.
    \item We conduct extensive experiments on a large-scale dataset. The results demonstrate that VulMorph-Fed substantially outperforms existing federated and centralised baselines in cross-project vulnerability detection, achieving superior privacy-utility trade-offs and scalability.
\end{itemize}

The remainder of this paper is structured as follows. Section \ref{sec:background} reviews related work in deep learning for vulnerability detection, cross-project prediction, and federated learning. Section \ref{sec:method} details the proposed VulMorph-Fed framework. Section \ref{sec:experiment} presents the experimental setup, while Section \ref{sec:results} discusses the results and ablation studies. Finally, Section \ref{sec:conclusion} concludes the paper.
