\section{Background and Related Work}
\label{sec:background}

\subsection{Deep Learning for Software Vulnerability Detection}
The application of Deep Learning (DL) to Software Vulnerability Detection (SVD) has evolved significantly from early sequence-based models to modern graph-based and transformer-based architectures. Early works, such as VulDeePecker \cite{li2018vuldeepecker} and SySeVR \cite{li2022sysevr}, treated code as sequences of tokens, employing recurrent neural networks (RNNs) like BiLSTMs to classify code gadgets. While these sequence-based models demonstrated the feasibility of DL for SVD, they struggled to capture the complex, non-linear structural dependencies inherent in source code.

The introduction of the Code Property Graph (CPG) \cite{yamaguchi2014modeling}, which unifies the Abstract Syntax Tree (AST), Control Flow Graph (CFG), and Data Flow Graph (DFG), marked a paradigm shift. Graph Neural Networks (GNNs) became the architecture of choice for SVD because they naturally align with the topological structure of CPGs. Devign \cite{zhou2019devign} pioneered this approach by applying Gated Graph Neural Networks (GGNNs) over composite graphs to detect vulnerabilities in large C/C++ projects. Subsequent GNN-based models have focused on improving graph representation and interpretability. For instance, DeepWukong \cite{cheng2021deepwukong} operates on inter-procedural program slices, while IVDetect \cite{li2021ivdetect} provides fine-grained interpretations using GNNExplainer. More recent advancements include Vul-LMGNNs \cite{sun2024vullmgnns}, which fuses Large Language Models (LLMs) with GNNs, and HAGNN \cite{toan2025detecting}, which utilises heterogeneous attention mechanisms.

Simultaneously, Transformer-based models and LLMs have been adapted for SVD. LineVul \cite{fu2022linevul} leverages CodeBERT for line-level vulnerability prediction, and LLMxCPG \cite{lekssays2025llmxcpg} uses CPGs to guide LLM attention. Despite these advances, recent studies \cite{weissberg2026llm, widyasari2026vultrial} indicate that LLMs often operate at a shallow structural level, highlighting the continued necessity for deep structural abstractions. Furthermore, rigorous replication studies \cite{chakraborty2022deep, daoudi2024revisiting, ding2025vulnerability} have consistently exposed a severe generalisation gap: SOTA models perform well in within-project settings but their accuracy drops precipitously when evaluated on unseen projects (cross-project).

\subsection{Cross-Project Vulnerability and Defect Prediction}
Cross-Project Vulnerability Detection (CPVD) and Cross-Project Defect Prediction (CPDP) aim to train models on a source project and deploy them on a target project. Traditional CPDP often relies on software metrics and transfer learning \cite{yamamoto2023flr}. In the context of DLVD, several works have employed Domain Adaptation (DA) techniques. CD-VulD \cite{liu2022cdvuld} and VulGDA \cite{li2023vulgda} use deep domain adaptation on graph embeddings to achieve cross-domain discovery. CPVD \cite{zhang2023cpvd} combines Graph Attention Networks (GATs) with adversarial domain adaptation, representing the current SOTA for fully supervised cross-project SVD. CSVD-TF \cite{tian2024csvdtf} fuses expert and semantic metrics using TrAdaBoost. However, a critical limitation of these DA-based approaches is that they require centralised access to raw source code from both the source and target domains, violating privacy constraints in many real-world scenarios.

\subsection{Federated Learning in Software Engineering}
Federated Learning (FL) \cite{mcmahan2017communication} allows multiple clients to jointly train a global model without sharing their local data. Its application to Software Engineering (FL-SE) is a burgeoning area of research, motivated by the proprietary nature of commercial codebases \cite{shanbhag2022exploring}. Recent empirical studies \cite{yang2024federated, hu2024vulfl} confirm the feasibility of FL for tasks like code clone detection and vulnerability detection.

However, adapting FL to SVD presents unique challenges, primarily due to severe non-IID data distributions (heterogeneity in CWE types and coding styles). FedGraphNN \cite{he2021fedgraphnn} demonstrated that standard federated GNNs often underperform centralised models on non-IID graph splits. Existing FL-SVD solutions typically rely on NLP models (e.g., CodeBERT) or naive GNN aggregations \cite{zhang2024vulnerability, hu2024vulfl}, which inherit the generalisation gap of their centralized counterparts. To address graph heterogeneity, approaches like FedDense \cite{wang2026feddense} decouple structural and feature learning. VulMorph-Fed builds upon these concepts but specifically targets the structural invariance of vulnerabilities by federating semantic prototypes rather than raw model parameters.

\subsection{Privacy-Preserving Mechanisms}
Ensuring data privacy in FL often involves Differential Privacy (DP) \cite{abadi2016deep} or Secure Aggregation \cite{bonawitz2017practical}. DP algorithms, such as DP-SGD, add calibrated noise (e.g., Laplace or Gaussian) to model updates. The magnitude of this noise is proportional to the sensitivity of the function being computed. Standard DP-SGD on large GNNs or LLMs requires significant noise due to the high dimensionality of the model parameters and the large vocabulary size of source code tokens, severely degrading utility. VulMorph-Fed mitigates this by applying DP at the level of semantic prototypes. By utilizing a compact, abstracted morphological space instead of the full token vocabulary, the sensitivity is drastically reduced, enabling a much more favorable privacy-utility trade-off.
