\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Resolving the Generalisation Crisis in Deep Learning-Based SVD}
The stark discrepancy in baseline performance across the eight datasets evaluated in Section \ref{sec:experiment} underscores the severity of the generalisation crisis in Software Vulnerability Detection. As noted by Daoudi et al. \cite{daoudi2024realvul} and Ding et al. \cite{ding2025primevul}, models that operate at the lexical or token level inevitably collapse when deployed on unseen projects. The failure of Centralised GAT and FedAvg+GAT on strict benchmarks like PrimeVul confirms that even advanced graph-based architectures are susceptible to lexical overfitting.

VulMorph-Fed provides a structural resolution to this crisis. By mathematically enforcing a many-to-one mapping $\Phi$ from concrete API tokens to an abstract semantic taxonomy $\mathcal{T}$, VCSA forces the GNN to learn the \emph{shape} of the control flow surrounding critical operations (e.g., unbounded memory copies, unchecked loop increments). Our results demonstrate that these structural morphologies are indeed invariant across projects, allowing VulMorph-Fed to maintain high F1-Scores across all eight disparate datasets without requiring access to the target domain's raw source code.

\subsection{Federated Prototypes vs. Parameter Averaging}
The ablation of the Morphology-Conditioned Federated Prototype Aggregation (MCFPA) module reveals the inadequacy of standard FedAvg for non-IID software engineering tasks. Source code exhibits extreme heterogeneity; client $A$ (e.g., a web server) and client $B$ (e.g., a kernel driver) will possess completely different distributions of CWE types. Averaging their GNN weight matrices indiscriminately dilutes the specialized knowledge encoded in each model. By federating prototypes $\mathbf{p}_c$ at the CWE class level and utilizing Morphology-Guided Message Passing (MGMP) to inject this knowledge locally via attention, VulMorph-Fed ensures that global intelligence directly augments the specific structural weaknesses present in any local codebase.

Furthermore, this prototype-based federation is the key to VulMorph-Fed's massive scalability (CCR $\approx 150$ KB) and high resilience to Differential Privacy noise ($\Delta f$ bounds). 

\subsection{Limitations and Threats to Validity}
Despite its strong cross-project performance, VulMorph-Fed has several limitations. First, the VCSA morphological abstraction relies on a finite, predefined taxonomy of 8 semantic types. While this taxonomy is sufficient to isolate a broad range of memory corruption and control flow vulnerabilities in C/C++, it may lack the granularity required to detect complex cryptographic misconfigurations or high-level business logic flaws. 

Second, the current implementation extracts proxy graphs directly from tokenized source code to avoid the enormous computational overhead of running full static analysis engines (e.g., Joern) at the edge. While these proxy graphs provide sufficient structural signal for our evaluation, integration with formally verified Code Property Graphs might further enhance precision.

Finally, while we simulate non-IID distributions by stratifying the cross-project splits, real-world industrial data may present unforeseen challenges. The distribution of CWE types in commercial codebases might be even more sparse or adversarial than simulated here. Future work will investigate dynamic, asynchronous prototype updates and Byzantine-robust aggregation mechanisms to handle extreme client heterogeneity and potential poisoning attacks.
